---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BGmisc

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R package version](https://www.r-pkg.org/badges/version/BGmisc)](https://cran.r-project.org/package=BGmisc)
[![Package downloads](https://cranlogs.r-pkg.org/badges/grand-total/BGmisc)](https://cran.r-project.org/package=BGmisc)</br>
[![status](https://joss.theoj.org/papers/ee3a025be4f61584f977a7657d936187/status.svg)](https://joss.theoj.org/papers/ee3a025be4f61584f977a7657d936187)

[![R-CMD-check](https://github.com/R-Computing-Lab/BGmisc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-Computing-Lab/BGmisc/actions/workflows/R-CMD-check.yaml)
[![Dev Main
branch](https://github.com/R-Computing-Lab/BGmisc/actions/workflows/R-CMD-dev_maincheck.yaml/badge.svg)](https://github.com/R-Computing-Lab/BGmisc/actions/workflows/R-CMD-dev_maincheck.yaml)
![License](https://img.shields.io/badge/License-GPL_v3-blue.svg)
<!-- badges: end -->

The BGmisc R package offers a comprehensive suite of functions tailored for extended behavior genetics analysis, including model identification, calculating relatedness, pedigree conversion, pedigree simulation, and more.

## Installation

You can install the released version of BGmisc from [CRAN](https://cran.r-project.org/) with:

```r
install.packages("BGmisc")
```

To install the development version of BGmisc from [GitHub](https://github.com/) use:

```r
# If devtools is not installed, uncomment the line below to install it before installing BGmisc.
# install.packages("devtools")
devtools::install_github("R-Computing-Lab/BGmisc")
```

## Citation
If you use BGmisc in your research or wish to refer to it, please cite the following paper:

```{r eval=TRUE, comment=NA}
citation(package = "BGmisc")
```


## Contributing
Contributions to the BGmisc project are welcome. For guidelines on how to contribute, please refer to the [Contributing Guidelines](https://github.com/R-Computing-Lab/BGmisc/blob/main/CONTRIBUTING.md). Issues and pull requests should be submitted on the GitHub repository. For support, please use the GitHub issues page.

## License
BGmisc is licensed under the GNU General Public License v3.0. For more details, see the [LICENSE.md](https://github.com/R-Computing-Lab/BGmisc/blob/main/LICENSE.md) file.

