---
title: "Harry Potter Pedigree"
output: rmarkdown::html_vignette
author: "Rachel Good"
vignette: >
  %\VignetteIndexEntry{Network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

This vignette showcases two key features that capitalize on the network
structure inherent in pedigrees:

1.  Finding extended families with *any* connecting relationships
    between members. This feature strictly uses a person's ID, mother's
    ID, and father's ID to find out which people in a dataset are
    remotely related by any path, effectively finding all separable
    extended families in a dataset.

2.  Using path tracing rules to quantify the *amount* of relatedness
    between all pairs of individuals in a dataset. The amount of
    relatedness can be characterized by additive nuclear DNA, shared
    mitochondrial DNA, sharing both parents, or being part of the same
    extended pedigree.

## Loading Required Libraries and Data

```{r setup}
library(BGmisc)
data(hp_ped)
```

# Finding Extended Families


```{r, echo=FALSE, results='hide', out.width='50%', fig.cap="hp_ped Pedigree"}
plotPedigree(hp_ped, code_male = 0, verbose = TRUE)
```

We will use the `hp_ped` pedigree data as an example.

```{r}
ds <- ped2fam(hp_ped, famID = "newFamID")
table(ds$FamID, ds$newFamID)
```

Because the `hp_ped` data already had a family ID variable we compare
our newly created variable to the pre-existing one. They match!

# Computing Relatedness

Once you know which sets of people are related at all to one another,
you'll likely want to know how much. For additive genetic relatedness,
you can use the `ped2add()` function.

```{r}
add <- ped2add(hp_ped)
```

This computes the additive genetic relatedness for everyone in the data.
It returns a square, symmetric matrix that has as many rows and columns
as there are IDs.

```{r}
add[1:7, 1:7]
```

The entry in the ith row and the jth column gives the relatedness
between person i and person j. For example, person 1 and person 11 share
`r add[1,11]`

```{r}
table(add)
```

It's probably fine to do this on the whole dataset when your data have
fewer than 10,000 people. When the data get large, however, it's much
more efficient to compute this relatedness separately for each extended
family.

```{r}
add_list <- lapply(
  unique(hp_ped$FamID),
  function(d) {
    tmp <- hp_ped[hp_ped$FamID %in% d, ]
    ped2add(tmp)
  }
)
```

## Other relatedness measures

The function works similarly for mitochondrial (`ped2mit`), common
nuclear environment through sharing both parents (`ped2cn`), and common
extended family environment (`ped2ce`).

### Computing mitochondrial relatedness

Here we calculate the mitochondrial relatedness between all pairs of
individuals in the `hp_ped` dataset.

```{r}
mit <- ped2mit(hp_ped)
mit[1:7, 1:7]
table(mit)
```

As you can see, some of the family members share mitochrondial DNA, such
as person 1 and person 3 `r mit[1,3]`, whereas person 1 and person 2 do
not.

### Computing relatedness through common nuclear environment

Here we calculate the relatedness between all pairs of individuals in
the `hp_ped` dataset through sharing both parents.

```{r}
commonNuclear <- ped2cn(hp_ped)
commonNuclear[1:7, 1:7]

table(commonNuclear)
```

### Computing relatedness through common extended family environment

Here we calculate the relatedness between all pairs of individuals in
the `hp_ped` dataset through sharing an extended family.

```{r}
extendedFamilyEnvironment <- ped2ce(hp_ped)
extendedFamilyEnvironment[1:7, 1:7]
table(extendedFamilyEnvironment)
```



# Tests
```{r}
plotPedigree(hp_ped, width = 1,code_male = "M",verbose = FALSE)

```

```{r}
data(inbreeding)
df <- inbreeding
summary(df)
plotPedigree(df, width = 1)

data(hazard)
df <- hazard
summary(df)
plotPedigree(df, width = 1)
```
