---
title: "Understanding and Computing Relatedness from Pedigree Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding and Computing Relatedness from Pedigree Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

When individuals share common ancestors, they are genetically related: they are expected to carry some proportion of alleles that are identical by descent. This expectation—called the relatedness coefficient—is central to many areas of genetics, including heritability estimation, pedigree-based modeling, twin and family studies, and the construction of kinship matrices for mixed-effects models.

Understanding relatedness is key for interpreting familial resemblance, controlling for shared genetic structure in statistical models, and simulating or analyzing traits across multigenerational pedigrees. But while the idea that “siblings are 50% related” is familiar, the reasoning behind such numbers—and how to compute them across complex family structures—is less transparent.

This vignette introduces the concept of relatedness from first principles and walks through how it is calculated from pedigree data. It begins with illustrative examples that explain expected relatedness values for familiar relationships using simplified functions. These examples clarify how shared ancestry translates into probabilistic expectations about genetic similarity.

From there, the vignette introduces a general-purpose matrix-based method for computing pairwise relatedness across pedigrees. Using the ped2com() function from the BGmisc package, we demonstrate how to build additive genetic relationship matrices under both complete and incomplete parentage, and we evaluate how different assumptions affect the resulting estimates. The goal is to provide a clear, rigorous, and practical guide to computing relatedness in real data.


# Relatedness Coefficient

The relatedness coefficient \( r \) indexes the proportion of alleles shared identically by descent (IBD) between two individuals. This value ranges from 0 (no shared alleles by descent) to 1 (a perfect genetic match, which occurs when comparing an individual to themselves, their identical twin, or their clone). Values can be interpreted in the context of standard relationships: e.g., full siblings are expected to have \( r = 0.5 \), half siblings \( r = 0.25 \), and first cousins \( r = 0.125 \).

Wright’s (1922) classic formulation computes 𝑟 by summing across shared ancestry paths:



\[
r_{bc} = \sum \left(\frac{1}{2}\right)^{n+n'+1} (1+f_a)
\]

Here, \( n \) and \( n' \) are the number of generations from each descendant to a common ancestor \( a \), and \( f_a \) is the inbreeding coefficient of \( a \), assumed to be zero unless specified otherwise.

The function `calculateRelatedness` computes the relatedness coefficient based on the number of generations back to common ancestors, whether the individuals are full siblings, and other parameters. The function can be used to calculate relatedness for various family structures, including full siblings, half siblings, and cousins.


```{r}
library(BGmisc)

# For full siblings, the relatedness coefficient is expected to be 0.5:
calculateRelatedness(generations = 1, full = TRUE)
# For half siblings, the relatedness coefficient is expected to be 0.25:
calculateRelatedness(generations = 1, full = FALSE)
```

The logic reflects the number and type of shared parents. For full siblings, both parents are shared (generational paths = 1 + 1), while for half siblings only one is (effectively halving the probability of sharing an allele). In otherwords, when `full = TRUE`, each sibling is one generation from the shared pair of parents, yielding `r=0.5`. When `full = FALSE`, they share only one parent, yielding `r=0.25`.

# Inferring r from Observed Phenotypic Correlation

In some cases, you observe a phenotypic correlation (e.g., height, cognition) between two individuals and want to infer what value of r would be consistent with that correlation under a fixed ACE model

The `inferRelatedness` function inverts the equation:

 \[
\text{obsR} = r \cdot a^2 + \text{sharedC} \cdot c^2
 \]

to solve for:

 \[
 r = \frac{\text{obsR} - \text{sharedC} \cdot c^2}{a^2}
 \]
 
where:
- `obsR` is the observed phenotypic correlation between two individuals or groups.
- `aceA` and `aceC` represent the proportions of variance due to additive genetic and shared environmental influences, respectively.
- `sharedC` is the shared-environment analog to the relatedness coefficient: it indicates what proportion of the shared environmental variance applies to this pair (e.g., 1 for siblings raised together, 0 for siblings raised apart).

```{r}
# Example usage:
# Infer the relatedness coefficient:
inferRelatedness(obsR = 0.5, aceA = 0.9, aceC = 0, sharedC = 0)
```


In this example, the observed correlation is 0.5, and no shared environmental variance is assumed. Given that additive genetic variance accounts for 90% of trait variance, the inferred relatedness coefficient is approximately 0.556. This reflects the proportion of genetic overlap that would be required to produce the observed similarity under these assumptions.


```{r}
# Now assume shared environment is fully shared:
inferRelatedness(obsR = 0.5, aceA = 0.45, aceC = 0.45, sharedC = 1)
```

In this case, the observed phenotypic correlation is still 0.5, and both additive genetic and shared environmental components are assumed to explain 45% of the variance. Because the shared environment is fully shared between individuals (sharedC = 1), much of the observed similarity is attributed to C, leaving only a small portion attributable to genetic relatedness. The function returns an inferred relatedness coefficient of approximately 0.11 — that is, the amount of additive genetic overlap required (under this model) to produce the remaining unexplained correlation after accounting for shared environmental similarity.
