---
title: "Network tools for finding extended pedigrees and path tracing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```


# Introduction

This vignette showcases two key features that capitalize on the network structure inherent in pedigrees:
1. Finding extended families with *any* connecting relationships between members. This feature strictly uses a person's ID, mother's ID, and father's ID to find out which people in a dataset are remotely related by any path, effectively finding all separable extended families in a dataset.
2. Using path tracing rules to quantify the *amount* of relatedness between all pairs of individuals in a dataset. The amount of relatedness can be characterized by additive nuclear DNA, shared mitochondrial DNA, sharing both parents, or being part of the same extended pedigree.


## Loading Required Libraries and Data

```{r setup}
library(BGmisc)
data(hazard)
```

# Finding Extended Families

Many pedigree datasets only contain information on the person, their mother, and their father, often without nuclear or extended family IDs.  Recognizing which sets of people are unrelated simplifies many pedigree-related tasks. This function facilitates those tasks by finding all the extended families.  People within the same extended family have at least some form of relation, however distant, while those in different extended families have no relations.

Using the `hazard` pedigree data as an example:

```{r}
ds <- ped2fam(hazard, famID = "newFamID")
table(ds$FamID, ds$newFamID)
```

Because the `hazard` data already had a family ID variable we compare our newly created variable to the pre-existing one.  They match!


# Computing Relatedness

Once you know which sets of people are related at all to one another, you'll likely want to know how much.  For additive genetic relatedness, you can use the `ped2add()` function.

```{r}
add <- ped2add(hazard)
```

This computes the additive genetic relatedness for everyone in the data.  It returns a square, symmetric matrix that has as many rows and columns as there are IDs.  

```{r}
add[1:7, 1:7]
```


The entry in the ith row and the jth column gives the relatedness between person i and person j. For example, person 1 and person 11 share  `r add[1,11]`

```{r}
table(add)
```

It's probably fine to do this on the whole dataset when your data have fewer than 10,000 people.  When the data get large, however, it's much more efficient to compute this relatedness separately for each extended family.

```{r}
add_list <- lapply(
  unique(hazard$FamID),
  function(d) {
    tmp <- hazard[hazard$FamID %in% d, ]
    ped2add(tmp)
  }
)
```

The function works similarly for mitochonrial (`ped2mit`), common nuclear environment through sharing both parents (`ped2cn`), and common extended family environment (`ped2ce`).

