---
title: "Trimming Large Pedigrees with BGmisc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trimming Large Pedigrees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

Large pedigrees collected in genetic or epidemiological studies often contain
many individuals who contribute little information to the analyses of interest.
Founders who appear as a parent of only one child, or terminal individuals with
no children and no recorded phenotype, can often be pruned from a dataset
without altering the connectivity or informational content of the core family
structure. This process --- *pedigree trimming* --- reduces computational
overhead for downstream analyses such as relatedness matrix estimation and
pedigree-based model fitting.

`BGmisc` provides two functions for this purpose:

- `findLeaves()`: identifies structural "leaf" nodes that are candidates for
  removal.
- `trimPedigree()`: iteratively removes those leaves, with optional stopping
  rules and phenotype-based protection.

Both functions operate on the directed pedigree graph produced internally by
`ped2graph()`, where edges run from **parent to child**. Node degree in this
representation has a direct pedigree interpretation:

| Degree | Pedigree meaning |
|---|---|
| indegree | number of recorded parents |
| outdegree | number of recorded children |

## Loading Required Libraries and Data

```{r setup}
library(BGmisc)
data(hazard)
data(inbreeding)
```

The `hazard` dataset contains `r nrow(hazard)` individuals across two extended
families and includes an `affected` status variable, making it well suited for
demonstrating phenotype-guided trimming. The `inbreeding` dataset contains
`r nrow(inbreeding)` individuals across eight families with various inbreeding
structures, useful for showing how trimming interacts with complex pedigree
topologies.

# Understanding Leaf Nodes

`findLeaves()` recognises two structural leaf types, both controlled by logical
flags.

**Terminal nodes** (`include_terminal = TRUE`, the default) are individuals with
no recorded children (outdegree 0). They sit at the outermost tips of the
pedigree. Removing them does not sever any path between the remaining
individuals.

**Founder singletons** (`include_founder_singletons = TRUE`, the default) are
founders --- individuals with no recorded parents (indegree 0) --- who appear as
a parent of exactly one child (outdegree 1). Because they connect to the network
through only a single edge, they can be removed without disconnecting any other
pair.

At least one of these types must be enabled; calling `findLeaves()` with both
disabled raises an error.

# Identifying Leaves with `findLeaves()`

## Basic usage

```{r}
leaves <- findLeaves(hazard)
leaves
```

There are `r length(leaves)` leaf candidates in the `hazard` pedigree. To
understand *why* each individual is flagged, use `verbose = TRUE`:

```{r, message = TRUE}
leaves <- findLeaves(hazard, verbose = TRUE)
```

## Controlling which leaf types to include

By default both leaf types are flagged. You can restrict to one or the other:

```{r}
# Terminal nodes only (outdegree == 0)
terminal_only   <- findLeaves(hazard, include_founder_singletons = FALSE)

# Founder singletons only (indegree == 0 & outdegree == 1)
singletons_only <- findLeaves(hazard, include_terminal = FALSE)

cat("Terminal only:    ", length(terminal_only),   "\n")
cat("Singletons only:  ", length(singletons_only),  "\n")
cat("Both combined:    ", length(leaves),            "\n")
```

Notice that the combined count can be smaller than the two individual counts
summed, because some individuals may qualify under both criteria simultaneously.

## Using a dataset with non-default column names

`findLeaves()` inherits the `personID`, `momID`, and `dadID` arguments from
`ped2fam()`. The `potter` dataset uses `personID` instead of `ID`:

```{r}
data(potter)
potter_leaves <- findLeaves(
  potter,
  personID = "personID"
)
length(potter_leaves)
```

# Iterative Trimming with `trimPedigree()`

`trimPedigree()` repeatedly calls `findLeaves()` and removes the identified
individuals, re-evaluating the pedigree structure after each pass. Trimming
peels the tree from the outside in: removing one layer of leaves may expose
individuals who were previously internal nodes and are now themselves leaves.

```{r, message = TRUE}
trimmed <- trimPedigree(hazard, verbose = TRUE)

cat("Before:", nrow(hazard), "individuals\n")
cat("After: ", nrow(trimmed), "individuals\n")
```

After trimming, `momID` and `dadID` columns are automatically updated so that
any reference to a removed individual is replaced with `NA`. You can verify
that no dangling references remain:

```{r}
removed_ids <- setdiff(as.character(hazard$ID), as.character(trimmed$ID))
sum(as.character(trimmed$dadID) %in% removed_ids, na.rm = TRUE)
sum(as.character(trimmed$momID) %in% removed_ids, na.rm = TRUE)
```

Both return 0, confirming the cleaned pedigree is self-consistent.

# Stopping Rules

Two parameters give fine-grained control over when iterative trimming halts
before all leaves are exhausted.

## `max_iter`: limit the number of passes

Setting `max_iter` trims only a fixed number of "shells" from the outside of the
pedigree. This preserves the broader structure while removing only the most
peripheral individuals.

```{r, message = TRUE}
trimmed_1 <- trimPedigree(hazard, max_iter = 1L, verbose = TRUE)
trimmed_2 <- trimPedigree(hazard, max_iter = 2L, verbose = TRUE)

cat("After 1 iteration: ", nrow(trimmed_1), "individuals\n")
cat("After 2 iterations:", nrow(trimmed_2), "individuals\n")
cat("Full trim:         ", nrow(trimmed),   "individuals\n")
```

## `min_size`: retain a minimum number of individuals

Setting `min_size` prevents the pedigree from being trimmed below a specified
number of rows. Trimming halts before any iteration that would violate this
threshold:

```{r, message = TRUE}
trimmed_min <- trimPedigree(hazard, min_size = 20L, verbose = TRUE)
cat("Retained:", nrow(trimmed_min), "individuals (minimum was 20)\n")
```

`max_iter` and `min_size` can be combined. Trimming stops as soon as either
condition is met:

```{r, message = TRUE}
trimmed_both <- trimPedigree(hazard, max_iter = 2L, min_size = 20L, verbose = TRUE)
cat("Retained:", nrow(trimmed_both), "individuals\n")
```

# Phenotype-Guided Trimming

In genetic epidemiology, the most important individuals in a pedigree are often
those with recorded phenotypic data. Removing such individuals discards valuable
information even when they are structural leaves. `findLeaves()` and
`trimPedigree()` both accept `keep_var` and `keep_vals` to protect
phenotypically informative individuals from removal at every iteration.

The `hazard` dataset includes an `affected` column that records disease status:

```{r}
table(hazard$affected, useNA = "ifany")
```

## Protect everyone with any recorded phenotype (`keep_var` only)

When `keep_vals` is omitted, any individual with a **non-missing** value in
`keep_var` is protected regardless of what that value is:

```{r, message = TRUE}
trimmed_any_pheno <- trimPedigree(hazard, keep_var = "affected", verbose = TRUE)

cat("Without phenotype protection:", nrow(trimmed),          "individuals\n")
cat("With phenotype protection:   ", nrow(trimmed_any_pheno), "individuals\n")
```

We can confirm that every individual with a recorded phenotype survived:

```{r}
protected_ids <- hazard$ID[!is.na(hazard$affected)]
all(protected_ids %in% trimmed_any_pheno$ID)
```

## Protect only individuals with a specific phenotype value

Use `keep_vals` when only certain phenotype values indicate that an individual
must be retained. Here, we protect only *affected* individuals (`TRUE`) and
allow removal of unaffected (`FALSE`) leaves:

```{r, message = TRUE}
trimmed_affected_only <- trimPedigree(
  hazard,
  keep_var  = "affected",
  keep_vals = TRUE,
  verbose   = TRUE
)
cat("Retained:", nrow(trimmed_affected_only), "individuals\n")

# Confirm all affected individuals are still present
affected_ids <- hazard$ID[isTRUE(hazard$affected) | hazard$affected == TRUE]
all(affected_ids %in% trimmed_affected_only$ID)
```

## Protect individuals with *missing* phenotype data

To retain individuals precisely because their phenotype is **not yet known**
(and may be collected in a follow-up wave), pass `keep_vals = NA`. This inverts
the default: only those with missing values are protected.

```{r, message = TRUE}
trimmed_keep_na <- trimPedigree(
  hazard,
  keep_var  = "affected",
  keep_vals = NA,
  verbose   = TRUE
)
cat("Retained:", nrow(trimmed_keep_na), "individuals\n")
```

## Combining protection criteria

`keep_vals` accepts a vector, so you can protect individuals with missing data
*and* those with a specific value simultaneously:

```{r, message = TRUE}
trimmed_na_or_true <- trimPedigree(
  hazard,
  keep_var  = "affected",
  keep_vals = c(TRUE, NA),
  verbose   = TRUE
)
cat("Retained:", nrow(trimmed_na_or_true), "individuals\n")
```

# Forced Removal with `remove_ids`

Before any leaf-based trimming, you can unconditionally remove specific
individuals using `remove_ids`. This is useful when particular individuals
should be excluded for reasons unrelated to pedigree topology --- for example,
data quality issues or consent withdrawal:

```{r, message = TRUE}
force_out <- head(hazard$ID, 2)

trimmed_forced <- trimPedigree(
  hazard,
  remove_ids = force_out,
  verbose    = TRUE
)
cat("IDs force-removed:", paste(force_out, collapse = ", "), "\n")
cat("Retained:", nrow(trimmed_forced), "individuals\n")
```

Forced removals are applied first; subsequent leaf passes then operate on the
already-reduced pedigree, which may expose new leaves that would not have
appeared otherwise.

# Summary

| Parameter | Controls |
|---|---|
| `include_terminal` | Whether outdegree-0 nodes are leaf candidates |
| `include_founder_singletons` | Whether indegree-0 & outdegree-1 nodes are leaf candidates |
| `max_iter` | Maximum number of trimming passes |
| `min_size` | Minimum pedigree size; trimming halts before breaching this threshold |
| `keep_var` | Phenotype column used for protection |
| `keep_vals` | Specific values (or `NA`) that shield individuals from removal |
| `remove_ids` | IDs to remove unconditionally before leaf-based trimming |

The two functions work together naturally. Use `findLeaves()` to *preview* which
individuals would be removed under a given set of parameters, then pass those
same parameters to `trimPedigree()` to carry out the removal iteratively:

```{r}
# Preview
candidates <- findLeaves(hazard, keep_var = "affected")
cat(length(candidates), "individuals are leaf candidates given phenotype protection\n")

# Execute
result <- trimPedigree(hazard, keep_var = "affected")
cat(nrow(hazard) - nrow(result), "individuals removed;", nrow(result), "retained\n")
```
