---
title: "Fitting Pedigree-Based Variance Component Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting Pedigree-Based Variance Component Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

This vignette extends the example from `vignette("v60_pedigree_model_fitting.Rmd", package = "BGmisc")` to show how to fit models to multiple families simultaneously. The key functions are `buildOneFamilyGroup()` and `buildPedigreeMx()`, which translate pedigree data into OpenMx model specifications.

# Scaling Up to Many Families


```{r krsp-prep, eval = run_models}
library(ggpedigree)  # for pedigree data)
library(tidyverse)
data("redsquirrels_full")

ped_krsp <- redsquirrels_full |>
  transmute(
    ID    = as.integer(personID),
    momID = as.integer(momID),
    dadID = as.integer(dadID),
    sex   = sex,
  famID = as.integer(famID),
  lrs = lrs
  ) 

cat("KRSP pedigree:", nrow(ped_krsp), "individuals,",
    n_distinct(ped_krsp$famID), "grids\n")
summarizeFamilies(ped_krsp, famID = "famID")$family_summary |> arrange(desc(count))
```

```{r, eval = FALSE && run_models}

minim_family_size <- 3

ped_krsp_subset <- ped_krsp |>
  group_by(famID) |>
  filter(sum(!is.na(lrs)) >= minim_family_size) |>
  ungroup()

n_families <- unique(ped_krsp_subset$famID)



# Pre-allocate storage
add_list     <- vector("list", length(n_families))
cn_list      <- vector("list", length(n_families))
obs_ids_list <- vector("list", length(n_families))
pheno_list   <- vector("list", length(n_families))

for (i in seq_along(n_families)) {
  
  ped_i <- subset(ped_krsp_subset, famID == n_families[i])
  phenotypic_ids <- ped_i$ID[!is.na(ped_i$lrs)]
  A_i  <- as.matrix(ped2add(ped_i, sparse = FALSE, keep_ids = phenotypic_ids))
  Cn_i <- as.matrix(ped2cn(ped_i,  sparse = FALSE, keep_ids = phenotypic_ids))
  
   n_i  <- nrow(A_i)
  id_order_i <- rownames(A_i)
  
  pheno_vals  <- ped_i$lrs[match(id_order_i, as.character(ped_i$ID))]
  
  obs_ids_i   <- make.names(id_order_i[!is.na(pheno_vals)])
  pheno_row_i <- matrix( as.double(pheno_vals[!is.na(pheno_vals)]), 
                         nrow = 1, 
                         dimnames = list(NULL, obs_ids_i))
  
  rownames(A_i) <- colnames(A_i) <- id_order_i
  
  rownames(Cn_i) <- colnames(Cn_i) <- id_order_i
    add_list[[i]]     <- A_i
  cn_list[[i]]      <- Cn_i
  obs_ids_list[[i]] <- obs_ids_i
  pheno_list[[i]]   <- pheno_row_i
}



group_models <- lapply(seq_len(n_families), function(i) {
  buildOneFamilyGroup(
    group_name  = paste0("ped", i),
    Addmat      = add_list[[i]],
    Nucmat      = cn_list[[i]],
    full_df_row = pheno_list[[i]],
    obs_ids     = obs_ids_list[[i]]
  )
})

multi_model <- buildPedigreeMx(
  model_name   = "MultiPedigreeModel",
  vars         = start_vars,
  group_models = group_models
)

fitted_multi <- mxRun(multi_model)
#saveRDS(fitted_multi, "fitted_multi.RDS")
```

---






# Part 3: Scaling Up — Many Families and Parameter Recovery

Once you understand the pattern from Parts 1 and 2, you can scale to many families using a loop. We switch to **simulated data** here so we have known true parameter values to check our estimates against.

## Simulate pedigrees and data

```{r simulate-many, eval = has_mvtnorm && run_models}
library(mvtnorm)
set.seed(2024)

# True variance components
true_var <- list(ad2 = 0.50, cn2 = 0.10, ee2 = 0.40,
                 ce2 = 0, mt2 = 0, dd2 = 0, am2 = 0)

n_families <- 10

# Pre-allocate storage
add_list     <- vector("list", n_families)
cn_list      <- vector("list", n_families)
obs_ids_list <- vector("list", n_families)
pheno_list   <- vector("list", n_families)

for (i in seq_len(n_families)) {
  ped_i <- simulatePedigree(kpc = 3, Ngen = 4, marR = 0.6)

  A_i  <- as.matrix(ped2add(ped_i, sparse = FALSE))
  Cn_i <- as.matrix(ped2cn(ped_i,  sparse = FALSE))
  n_i  <- nrow(A_i)

  # Implied covariance from true parameters
  V_i <- true_var$ad2 * A_i +
         true_var$cn2 * Cn_i +
         true_var$ee2 * diag(1, n_i)

  y_i <- rmvnorm(1, sigma = V_i)

  ids_i <- make.names(rownames(A_i))
  rownames(A_i) <- colnames(A_i) <- ids_i
  rownames(Cn_i) <- colnames(Cn_i) <- ids_i

  add_list[[i]]     <- A_i
  cn_list[[i]]      <- Cn_i
  obs_ids_list[[i]] <- ids_i
  pheno_list[[i]]   <- matrix(y_i, nrow = 1, dimnames = list(NULL, ids_i))
}

cat("Family sizes:", vapply(add_list, nrow, integer(1)), "\n")
```

## Build and fit the multi-family model

```{r fit-many, eval = has_mvtnorm && run_models}
group_models <- lapply(seq_len(n_families), function(i) {
  buildOneFamilyGroup(
    group_name  = paste0("ped", i),
    Addmat      = add_list[[i]],
    Nucmat      = cn_list[[i]],
    full_df_row = pheno_list[[i]],
    obs_ids     = obs_ids_list[[i]]
  )
})

multi_model <- buildPedigreeMx(
  model_name   = "MultiPedigreeModel",
  vars         = start_vars,
  group_models = group_models
)

fitted_multi <- mxRun(multi_model)
```

## Compare estimates to true values

```{r compare-many, eval = has_mvtnorm && run_models}
data.frame(
  Component = c("Additive genetic (Vad)",
                "Common nuclear  (Vcn)",
                "Unique environ. (Ver)"),
  True      = c(true_var$ad2, true_var$cn2, true_var$ee2),
  Estimated = round(c(
    fitted_multi$ModelOne$Vad$values,
    fitted_multi$ModelOne$Vcn$values,
    fitted_multi$ModelOne$Ver$values
  ), 4)
)
```

With ten families, estimates should be substantially closer to the true values than with a single family.

## Using the high-level wrapper

For convenience, `fitPedigreeModel()` wraps the build and fit steps together, and uses `mxTryHard()` for robust optimization:

```{r fit-wrapper, eval = has_mvtnorm && run_models}
fitted_easy <- fitPedigreeModel(
  model_name   = "EasyFit",
  vars         = start_vars,
  group_models = group_models,
  tryhard      = TRUE
)
summary(fitted_easy)
```

---

# Understanding the Covariance Structure

The model estimated above is defined by:

$$V = \sigma^2_a \mathbf{A} + \sigma^2_{cn} \mathbf{C}_n + \sigma^2_e \mathbf{I}$$

where $\mathbf{A}$ is the additive genetic relatedness matrix, $\mathbf{C}_n$ is the common nuclear environment matrix, $\mathbf{I}$ is the identity matrix (unique environment), and the $\sigma^2$ terms are the variance components estimated by the model.

This is an extension of the classical twin model to arbitrary pedigree structures. Additional components can be added by passing `Extmat` (common extended environment), `Mtdmat` (mitochondrial), or `Dmgmat` (dominance) to `buildOneFamilyGroup()`. See `vignette("v1_modelingvariancecomponents", package = "BGmisc")` for background on identification.

---

# Summary

| Part | Data | Key point |
|------|------|-----------|
| 1 — Single family | `hazard`, family 1 | Full pipeline, one group model |
| 2 — Two families | `hazard`, families 1 & 2 | Same steps twice; `buildPedigreeMx()` combines them |
| 3 — Many families | Simulated | Loop over families; verify parameter recovery |

The helper functions (`buildOneFamilyGroup()`, `buildPedigreeMx()`, `fitPedigreeModel()`) handle the translation from pedigree relatedness matrices into OpenMx model specifications. The pattern is always the same: one call to `buildOneFamilyGroup()` per family, then one call to `buildPedigreeMx()` to combine them.
