---
title: "Extended: Fitting Pedigree-Based Variance Component Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extended: Fitting Pedigree-Based Variance Component Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
#source('https://vipbg.vcu.edu/vipbg/OpenMx2/software/getOpenMx.R')

library(BGmisc)

has_openmx <- requireNamespace("OpenMx", quietly = TRUE)

if (has_openmx) {
  library(OpenMx)
  mxOption(key="Number of Threads", value=imxGetNumThreads())

} else {
  message(
    "OpenMx is not installed. Code will be shown but not executed.\n",
    "Install with: install.packages('OpenMx')"
  )
}

bgmisc_testing_env <- Sys.getenv("BGMISC_TESTING", unset = "")
bgmisc_testing <- tolower(bgmisc_testing_env) %in% c("1", "true", "yes")
run_models <- has_openmx && interactive() #&& !bgmisc_testing
rds_dir <- file.path( "inst/extdata/")
rds_file_ace <- file.path(rds_dir, "fitted_multi_ace.rds")
rds_file_mace <- file.path(rds_dir, "fitted_multi_mace.rds")
```

# Introduction

This vignette extends the example from `vignette("v60_pedigree_model_fitting", package = "BGmisc")` to show how to fit models to multiple families simultaneously. The key functions are `buildOneFamilyGroup()` and `buildPedigreeMx()`, which translate pedigree data into OpenMx model specifications.

# Scaling Up to Many Families


Here we replicate several estimates of heritability across multiple families of red squirrels.  We use the `redsquirrels_full` dataset from the `ggpedigree` package, which contains pedigree and phenotypic data on red squirrels from the Kluane region of the Yukon, Canada.  The phenotype we analyze is lifetime reproductive success (LRS), which is a count of the number of offspring that survive to a certain age.

```{r krsp-prep}
library(ggpedigree) # for pedigree data)
library(tidyverse)
data("redsquirrels_full")

ped_krsp <- redsquirrels_full |>
  transmute(
    ID = as.integer(personID),
    momID = as.integer(momID),
    dadID = as.integer(dadID),
    sex = sex,
    famID = as.integer(famID),
    lrs = lrs
  )

cat(
  "KRSP pedigree:", nrow(ped_krsp), "individuals,",
  n_distinct(ped_krsp$famID), "grids\n"
)
summarizeFamilies(ped_krsp, famID = "famID")$family_summary |> arrange(desc(count))
```

```{r}
minim_family_size <- 10

ped_krsp_subset <- ped_krsp |>
  group_by(famID) |>
  filter(sum(!is.na(lrs)) >= minim_family_size) |>
  ungroup()

id_families <- unique(ped_krsp_subset$famID)
n_families <- length(id_families)

# Pre-allocate storage
add_list <- vector("list", length(n_families))
cn_list <- vector("list", length(n_families))
mt_list <- vector("list", length(n_families))
obs_ids_list <- vector("list", length(n_families))
pheno_list <- vector("list", length(n_families))


# Starting values for variance components
start_vars <- list(
  ad2 = 0.3, # additive genetic
  cn2 = 0.1, # common nuclear environment
  ce2 = 0, # common extended (not estimated here)
  mt2 = 0.1, # mitochondrial
  dd2 = 0, # dominance (not estimated here)
  am2 = 0, # A x Mt interaction (not estimated here)
  ee2 = 0.5 # unique environment
)
```
```{r, eval = has_openmx}
for (i in seq_len(n_families)) {
  ped_i <- subset(ped_krsp_subset, famID == id_families[i])
  phenotypic_ids <- ped_i$ID[!is.na(ped_i$lrs)]
  A_i <- as.matrix(ped2add(ped_i, sparse = FALSE, keep_ids = phenotypic_ids))
  Cn_i <- as.matrix(ped2cn(ped_i, sparse = FALSE, keep_ids = phenotypic_ids))
  Mt_i <- as.matrix(ped2mit(ped_i, sparse = FALSE, keep_ids = phenotypic_ids))

  n_i <- nrow(A_i)
  id_order_i <- rownames(A_i)

  pheno_vals <- ped_i$lrs[match(id_order_i, as.character(ped_i$ID))]

  obs_ids_i <- make.names(id_order_i[!is.na(pheno_vals)])
  pheno_row_i <- matrix(as.double(pheno_vals[!is.na(pheno_vals)]),
    nrow = 1,
    dimnames = list(NULL, obs_ids_i)
  )

  rownames(A_i) <- colnames(A_i) <- obs_ids_i
  rownames(Cn_i) <- colnames(Cn_i) <- obs_ids_i
  rownames(Mt_i) <- colnames(Mt_i) <- obs_ids_i
  
  add_list[[i]] <- A_i
  cn_list[[i]] <- Cn_i
  mt_list[[i]] <- Mt_i
  obs_ids_list[[i]] <- obs_ids_i
  pheno_list[[i]] <- pheno_row_i
}


group_models_ace <- lapply(seq_len(n_families), function(i) {
  buildOneFamilyGroup(
    group_name  = paste0("ped", i),
    Addmat      = add_list[[i]],
    Nucmat      = cn_list[[i]],
    Mtdmat      = NULL,
    full_df_row = pheno_list[[i]],
    obs_ids     = obs_ids_list[[i]]
  )
})

group_models_mace <- lapply(seq_len(n_families), function(i) {
  buildOneFamilyGroup(
    group_name  = paste0("ped", i),
    Addmat      = add_list[[i]],
    Nucmat      = cn_list[[i]],
    Mtdmat      = mt_list[[i]],
    full_df_row = pheno_list[[i]],
    obs_ids     = obs_ids_list[[i]]
  )
})


multi_model_ace <- buildPedigreeMx(
  model_name   = "MultiPedigreeModel",
  vars         = start_vars,
  group_models = group_models_ace,
  ci = TRUE
)

multi_model_mace <- buildPedigreeMx(
  model_name   = "MultiPedigreeModel",
  vars         = start_vars,
  group_models = group_models_mace,
  ci = TRUE
)
```
```{r, include = FALSE}
fitted_multi_ace <- NULL # ensure variable exists for later use
fitted_multi_mace <- NULL # ensure variable exists for later use
```
```{r, eval = run_models}
fitted_multi_ace <- mxRun(multi_model_ace,
                          intervals= TRUE)
saveRDS(fitted_multi_ace, "inst/extdata/fitted_multi_ace.rds")

fitted_multi_mace <- mxRun(multi_model_mace, intervals= TRUE)
saveRDS(fitted_multi_mace, "inst/extdata/fitted_multi_mace.rds")
```

```
fitted_multi_ace <- mxRun(multi_model_ace)
fitted_multi_mace <- mxRun(multi_model_mace)
```
```{r, include = FALSE, eval = file.exists(rds_file_ace) && !run_models && file.exists(rds_file_mace)}
fitted_multi_ace <- readRDS(rds_file_ace)
fitted_multi_mace <- readRDS(rds_file_mace)
```
```{r, eval = has_openmx && !is.null(fitted_multi_ace)}
summary(fitted_multi_ace)
summary(fitted_multi_ace)$CI
total_var_ace <- sum(fitted_multi_ace$ModelOne$Vad$values, 
                 fitted_multi_ace$ModelOne$Vcn$values,
                 #fitted_multi_ace$ModelOne$Vmt$values, 
                 fitted_multi_ace$ModelOne$Ver$values)
```


```{r total-fam-results, eval = has_openmx && !is.null(fitted_multi_ace)}
cat("Additive genetic (Vad):", fitted_multi_ace$ModelOne$Vad$values/total_var_ace, "\n")
cat("Common nuclear  (Vcn):", fitted_multi_ace$ModelOne$Vcn$values/total_var_ace, "\n")
cat("Unique environ. (Ver):", fitted_multi_ace$ModelOne$Ver$values/total_var_ace, "\n")
```

As you can see, we have fit a multigroup pedigree model using `r n_families` families of several thousand squirrels. The model includes additive genetic variance (Vad), common nuclear environmental variance (Vcn), and unique environmental variance (Ver). The mitochondrial variance component (Vmt) was included in the MACE model but not estimated in the ACE model. The results show the proportion of total variance attributed to each component, which can be interpreted as heritability and environmental contributions to the phenotype of interest (LRS) in these
squirrels.

```{r, eval = has_openmx &&  !is.null(fitted_multi_mace)}
summary(fitted_multi_mace)
summary(fitted_multi_mace)$CI
total_var_mace <- sum(fitted_multi_mace$ModelOne$Vad$values, 
                 fitted_multi_mace$ModelOne$Vcn$values,
                 fitted_multi_mace$ModelOne$Vmt$values, 
                 fitted_multi_mace$ModelOne$Ver$values)
```


```{r total-fam-results-mace, eval = has_openmx && !is.null(fitted_multi_ace)}
cat("Additive genetic (Vad):", fitted_multi_mace$ModelOne$Vad$values/total_var_mace, "\n")
cat("Common nuclear  (Vcn):", fitted_multi_mace$ModelOne$Vcn$values/total_var_mace, "\n")
cat("Mitochondrial (Vmt):", fitted_multi_mace$ModelOne$Vmt$values/total_var_mace, "\n")
cat("Unique environ. (Ver):", fitted_multi_mace$ModelOne$Ver$values/total_var_mace, "\n")
```
Now we can compare the ACE and MACE models to see if including the mitochondrial component changes our estimates of heritability and environmental contributions. This can provide insights into the role of mitochondrial inheritance in the trait of interest (LRS) in these squirrels.

```{r, eval = has_openmx && !is.null(fitted_multi_ace) && !is.null(fitted_multi_mace)}

mxCompare(fitted_multi_mace, fitted_multi_ace)

fitted_multi_ace_ci <- omxRunCI(fitted_multi_ace, verbose = 0, optimizer = "SLSQP")

```
